\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-02-17 misc/JSS article/jss-article-rnw/article.tex

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++ based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

@@

2   4  1  0    21.5  22.10    23.02      23.17     24.4  24.38    24.4
3   4  1  1    21.4  22.29    27.74      27.85     30.4  31.79    33.9

@@

1  9.964 0.023670
2 14.045 0.006743

@@

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, can handle complex labelled data without loss of information, and offers best-in-\proglang{R} computational performance and memory efficiency. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
      how = "longer", na.rm = FALSE, factor = c("names", "labels"),
      check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
      sort = FALSE, transpose = FALSE)
\end{Code}
%
Below, its use is demonstrated with a generated dataset about fruits. We could also think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples with real data.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+              type_name = rep(c("Apples", "Bananas"), each = 2),
+              id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id      r      h
1    A    Apples  1 0.1054 3.5624
2    A    Apples  2 0.7924 3.9380
3    B   Bananas  1 0.1717 1.1828
4    B   Bananas  2 0.3461 0.3884
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named 'lab\_name'. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the 'variable' and 'value' columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label  value
1    A    Apples  1        r Fruit Radius 0.1054
2    A    Apples  2        r Fruit Radius 0.7924
3    B   Bananas  1        r Fruit Radius 0.1717
4    B   Bananas  2        r Fruit Radius 0.3461
5    A    Apples  1        h Fruit Height 3.5624
6    A    Apples  2        h Fruit Height 3.9380
7    B   Bananas  1        h Fruit Height 1.1828
8    B   Bananas  2        h Fruit Height 0.3884
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
In general, \fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id    r_A    r_B   h_A    h_B
1  1 0.1054 0.1717 3.562 1.1828
2  2 0.7924 0.3461 3.938 0.3884
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named 'variable' exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's 'from' or 'to' elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A      B
1  1        r Fruit Radius 0.1054 0.1717
2  2        r Fruit Radius 0.7924 0.3461
3  1        h Fruit Height 3.5624 1.1828
4  2        h Fruit Height 3.9380 0.3884
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without IDs resembles a generalization of \code{data.table::transpose}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often, in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{List Processing Functions} attempt to provide a basic toolkit. \newline

@@

Quantiles
          1%     5%    10%    25%    50%    75%    90%    95%    99%
FALSE  41.32  37.63  48.98   57.5  65.87  69.68   74.1  32.39  76.18
TRUE   56.67  65.65  69.69  71.84  75.32  78.61  81.26  81.23  83.59
-----------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}

@@

This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. All packages are compiled and used in single-threaded mode to avoid artifacts of different parallel setups and resource utilization. The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets.

@@

R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)

@@

+                 Rfast = Rfast::colmeans(m),

@@

        R 9.489 9.538 9.622  9.587 9.625 11.013   100
    Rfast 4.732 4.757 4.974  4.779 4.826 22.022   100
 collapse 1.356 1.459 1.685  1.564 1.737  7.917   100

@@

          R 10.721 11.036 12.69 11.323 11.901 51.54   100
 data.table 16.200 16.695 18.56 16.940 18.126 60.54   100
   collapse  5.109  5.512  6.08  5.617  5.952 16.85   100

@@

 data.table 130.1 130.9 132.5  131.8 133.1 150.5   100
   collapse 117.1 118.1 119.5  118.6 119.6 130.3   100

@@

 data.table 21.136 23.775 34.691 25.463 45.039 72.758   100
   collapse  1.178  1.323  1.348  1.356  1.377  1.509   100

@@

 data.table 11.87 14.86 19.92  16.07 17.60 56.57   100
   collapse 11.67 12.69 19.28  15.83 17.27 60.33   100

@@

 data.table 63.45 102.5 104.88 104.89 108.03 140.38   100
   collapse 33.23  71.4  74.14  72.77  75.69  93.31   100

@@

      base_int 60.165 62.240 64.93 64.143 65.25 113.92   100
  collapse_int  8.518  9.183 11.16  9.374 12.51  54.94   100
     base_char 92.549 95.915 98.02 97.444 99.13 124.71   100
 collapse_char 22.018 23.435 24.55 23.829 24.36  34.54   100

@@

        base_int 27.159 27.309 29.033 28.026 30.473 32.37    10
    collapse_int  8.592  8.626  9.294  8.808  9.155 13.23    10
       base_char 76.203 76.389 80.089 77.340 82.390 88.61    10
 data.table_char 41.366 41.470 42.315 41.750 42.040 47.57    10
   collapse_char 26.039 26.153 26.871 26.369 26.630 31.65    10

>>>>

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

@@

2   4  1  0    21.5  21.90    23.02      23.16     24.4  24.38    24.4
3   4  1  1    21.4  22.37    27.74      28.28     30.4  31.51    33.9

@@

1  8.885 0.023614
2 13.685 0.006724

@@

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, can handle complex labelled data without loss of information, and offers best-in-\proglang{R} computational performance and memory efficiency. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
      how = "longer", na.rm = FALSE, factor = c("names", "labels"),
      check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
      sort = FALSE, transpose = FALSE)
\end{Code}
%
Below, its use is demonstrated with a generated dataset about fruits. We could also think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples with real data.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+              type_name = rep(c("Apples", "Bananas"), each = 2),
+              id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id      r      h
1    A    Apples  1 0.9322 0.4315
2    A    Apples  2 0.2230 0.7225
3    B   Bananas  1 0.8147 0.4918
4    B   Bananas  2 0.5642 2.4131
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named 'lab\_name'. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the 'variable' and 'value' columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label  value
1    A    Apples  1        r Fruit Radius 0.9322
2    A    Apples  2        r Fruit Radius 0.2230
3    B   Bananas  1        r Fruit Radius 0.8147
4    B   Bananas  2        r Fruit Radius 0.5642
5    A    Apples  1        h Fruit Height 0.4315
6    A    Apples  2        h Fruit Height 0.7225
7    B   Bananas  1        h Fruit Height 0.4918
8    B   Bananas  2        h Fruit Height 2.4131
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
In general, \fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id    r_A    r_B    h_A    h_B
1  1 0.9322 0.8147 0.4315 0.4918
2  2 0.2230 0.5642 0.7225 2.4131
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named 'variable' exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's 'from' or 'to' elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A      B
1  1        r Fruit Radius 0.9322 0.8147
2  2        r Fruit Radius 0.2230 0.5642
3  1        h Fruit Height 0.4315 0.4918
4  2        h Fruit Height 0.7225 2.4131
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without IDs resembles a generalization of \code{data.table::transpose}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often, in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{List Processing Functions} attempt to provide a basic toolkit. \newline

@@

Quantiles
          1%     5%    10%    25%    50%    75%    90%    95%    99%
FALSE  41.39  45.78  49.08  57.51  65.98  70.14  74.12  75.63  76.91
TRUE   56.65  65.98   69.7  71.85  75.38  78.64  81.26  82.43   83.6
-----------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}

@@

This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.

@@

R> setDTthreads(4)
R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)

@@

+                 Rfast = Rfast::colmeans(m, parallel = TRUE, cores = 4),

@@

        R 9.632 9.837 9.959  9.852 9.940 12.70   100
    Rfast 1.315 1.805 2.257  1.840 1.917 11.92   100
 collapse 1.321 1.432 1.847  1.506 1.603 16.33   100

@@

          R 10.715 11.19 11.550 11.295 11.61 16.89   100
 data.table  9.154 18.70 19.743 19.229 20.19 30.22   100
   collapse  2.330  2.89  3.619  3.078  3.71 15.94   100

@@

 data.table 136.36 137.50 140.23 138.54 140.62 173.7   100
   collapse  68.42  76.85  82.25  80.28  83.36 163.3   100

@@

 data.table 8.915 10.881 16.398 13.166 16.903 73.045   100
   collapse 1.228  1.379  1.452  1.413  1.459  2.337   100

@@

 data.table 11.79 12.84 17.99  15.87 17.68 73.75   100
   collapse 11.73 13.77 20.37  15.69 18.17 76.75   100

@@

 data.table 60.01 69.99 90.12  74.28 111.73 157.7   100
   collapse 32.59 41.80 63.15  48.87  63.98 177.0   100

@@

      base_int 60.07 61.852  67.56 65.613  68.55 133.05   100
  collapse_int  6.69  9.222  10.34  9.581  10.18  17.56   100
     base_char 91.79 94.570 102.38 99.497 103.65 170.41   100
 collapse_char 22.29 23.879  26.83 24.577  27.58  80.46   100

@@

        base_int 26.873 26.993 28.261 27.094 27.306 33.25    10
    collapse_int  8.603  8.633  9.352  8.662  8.723 15.53    10
       base_char 84.091 84.726 86.127 85.100 86.008 91.23    10
 data.table_char 40.970 41.169 42.597 41.307 41.407 48.49    10
   collapse_char 30.482 30.541 30.774 30.622 30.803 31.47    10
